Deep mutational scanning of hemagglutinin helps identify evolutionarily successful human H3N2 influenza viruses

### 2018-02-01 paper/manuscript.tex
updated author contributions
final revisions

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,d,e,1]{Juhye M. Lee}
\author[b,f,1]{John Huddleston} 
\author[a,d,e]{Michael B. Doud}
\author[a,f]{Kathryn A. Hooper}
\author[b,c,2]{Trevor Bedford}
\author[a,c,d,2]{Jesse D. Bloom}

@@

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{J.M.L. performed the experiments; J.M.L. and J.H. analyzed data with input from T.B. and J.D.B.; J.M.L. and J.D.B. wrote the paper with input from J.H. and T.B.; all authors designed and planned the project}
\authordeclaration{The authors declare no conflict of interest.}
\equalauthors{\textsuperscript{1}J.M.L. and J.H. contributed equally to this work.}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed: trevor@bedford.io, jbloom@fredhutch.org}

@@

Two main characteristics distinguish the evolutionarily successful trunk from its competitors: greater antigenic change, and efficient viral growth and transmission.
In principle, experiments could be informative for identifying how mutations affect these features.
Most work on influenza evolution to date has utilized experimental data primarily to assess the antigenicity of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction}.
While the antigenic features of a virus certainly contribute to its success, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
Specifically, due to influenza virus's high mutation rate~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and lack of intra-segment recombination~\cite{boni2008homologous}, deleterious mutations become linked to beneficial ones.
The resulting accumulation of deleterious mutations can affect non-antigenic properties that are central to viral fitness~\cite{luksza2014predictive}.
However, while extensive work has gone into characterizing how mutations affect HA antigenicity~\cite{laver1979antigenic, webster1980determination,wiley1981structural,koel2013substitutions,chambers2015identification,li2016selection}, we lack any large-scale quantitative characterization of how mutations to H3N2 HA affect non-antigenic properties crucial to viral fitness.

It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of large numbers of mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high}.
However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
Here we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain, A/Perth/16/2009.
We then examine whether these measurements are useful for distinguishing between H3N2 trunk and side-branch lineages, and find that mutations on the trunk are measured to be more favorable than those on the side branches, both at epitope and non-epitope sites.
However, the experiments are useful for assessing the effects of mutations only in closely related viral strains, since we show that mutations often have disparate effects on H3 and H1 HAs.
Our work highlights the potential for using high-throughput experimental measurements of mutational effects to inform evolutionary forecasting of human seasonal influenza virus.

@@

We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates. 
We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
Because there are 566 residues in HA, there are $566 \times 20 = 11,320$ estimated amino-acid preferences, corresponding to $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
The biological replicates are fairly well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
Replicate 1 exhibited the weakest correlation with the other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity during viral passage contributes most of the experimental noise.

\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
This result indicates that our experiments authentically capture some of the constraints on HA evolution. 
The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments (the $\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for by the constraints measured in the deep mutational scanning.
The ExpCM stringency parameter~\cite{hilton2017phydms} is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.

@@

A closer examination of Figure~\ref{fig:logoplot} reveals that the experimentally measured amino-acid preferences generally agree with existing knowledge about HA's structure and function.
For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} possess high preference for cysteine.
At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such as Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change,yang2015structure}.
A positively charged amino acid at site 329 is important for cleavage of the HA0 precursor into the mature form~\cite{stech2005new}, and this site strongly prefers arginine.
However, a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine. 
This codon is part of the signal peptide and is cleaved from the mature HA protein.
One possible reason that our experiments do not show a strong preference for methionine at this site could be alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.
% No upstream or downstream "ATG"'s are present. I did find an in-frame "ATC" one codon upstream of the canonical start site (in the non-coding region), and interestingly, all Perth/2009 HA sequences deposited in GenBank have an "ATT" in this non-coding position, including Seema's Perth HA.
% I6M(HA2) mutant slightly increases the fusion pH (Daniels 1985 Cell)
% We see high preference for T,S,Q over the wildtype Ile at site 226. Although Q226 in an H3 background is characterized as having preference for avian-type receptors, most of these studies were done in older H3 strains such as Aichi. Wu 2017 Cell Host Microbe has shown extensive epistasis in the RBS, specifically in the 220-loop, so perhaps the amino acids tolerated in this loop have shifted over time. Indeed the wildtype amino acids at 225 and 226 are different in Perth compared to HK68 and Aichi.

@@

\subsection*{There is less difference in mutational tolerance between the HA head and stalk domains for H3 than for H1}
Our experiments measure which amino acids are tolerated at each HA site under selection for protein function.
We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we quantify as the Shannon entropy of the re-scaled preferences.
In prior mutational studies of the WSN/1933 H1 HA, the stalk domain was found to be substantially less mutationally tolerant than the globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.

We performed a similar analysis using the current data for the Perth/2009 H3 HA.
Surprisingly, there was much less contrast in mutational tolerance between the stalk and head domains for the H3 HA than for the H1 (Figure~\ref{fig:mut_tolerance}).
For instance, in the H3 HA, the short helix A in the stalk is very mutationally tolerant.
Interestingly, there are more reports of readily selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2017quantifying,anderson2017natural} HAs.

@@

To investigate whether our experiments can aid in distinguishing trunk and side-branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the H3N2 phylogeny (Figure~\ref{fig:H3N2_phylogeny}).
The mutations that occurred on the trunk of the tree were consistently more beneficial for viral growth according to our experimental measurements (Figure~\ref{fig:trunkvssidebranch}A).
This difference in the effects of mutations between the trunk and side-branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
Some influenza sequences are derived from egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{mcwhite2016sequence}.
These mutations mostly occur on the terminal side branches that lead to tip nodes on the phylogenetic tree.
We therefore re-calculated the statistics separating side branches to tip and internal nodes, and again found that the difference between the trunk and both sets of side branches was statistically significant (Figure~\ref{fig:trunkvssidebranch}B).
Therefore, strains with mutations that we experimentally measure to be favorable for viral growth tend to do better in nature than strains with mutations that we measure to be less favorable.

@@

Mutations that occur on the trunk are scored as significantly more favorable in our experiments than side-branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}A).
Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and at sites that are also under immune pressure.
The fact that our experiments can discriminate between the trunk and side branches at epitope sites despite only assaying for viral growth underscores the fact that mutations at epitope sites still have significant non-antigenic functional consequences~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.

However, there are differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
The per-site sequence preference at epitope sites increases over time, whereas at non-epitope sites it remains relatively constant  (Figure~\ref{fig:sequence_preference}B).
%In addition, the per-site sequence preference is consistently higher at non-epitope than epitope sites, perhaps because immune pressure sometimes drives the fixation of functionally less favorable mutations at epitope sites.
The fact that only epitope sites exhibit an increase in sequence preference over time can also be rationalized in terms of prior knowledge about protein evolution.
As proteins evolve, there can be epistatic shifts in the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some epitope sites in H3 HA~\cite{wu2017diversity}.
It is also known that immune pressure can increase epistasis among substitutions by selecting for combinations of functionally deleterious immune-escape mutations and counterbalancing secondary mutations~\cite{gong2014epistatically}.
We hypothesize that immune pressure increases the accumulation of epistatic interactions involving substitutions at epitope sites.
Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA (Figure~\ref{fig:sequence_preference}B).
Such a trend is not apparent for non-epitope sites, perhaps because they are under less selection to fix adaptive mutations.

@@

An obvious hypothesis for why the H1 deep mutational scanning is not informative about the evolutionary fate of H3 strains is that the effect of the same mutation will often be very different between these two HA subtypes.
To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2017mapping}.
However, the H1 and H3 HAs are far more diverged, with only 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
One simple way to investigate the extent of shifts in amino-acid preferences is to correlate measurements from independent deep mutational scanning replicates on H1 and H3 HAs.
Figure~\ref{fig:distance_distribution}B shows that replicate measurements on the same HA variant are more correlated than those on different HA variants.

@@

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=12cm]{figs/distance_distribution/distance_distribution.pdf}
\caption{\label{fig:distance_distribution}
{\bf There are substantial differences in the effects of mutations between H1 and H3 HAs.}
(A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled. 
These HAs have 42\% amino-acid identity.
(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study  and the three replicates in prior deep mutational scanning of an H1 HA~\cite{doud2016accurate}.
Comparisons between H3 replicates are in purple, those between H1 replicates are in brown, and those across H1 and H3 replicates are in gray. 
$R$ indicates the Pearson correlation coefficient.
(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2017mapping}, and plotted the distribution of shifts for all sites.
The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
The shifts are also much larger than those previously observed between two variants of HIV envelope protein (Env) that share 86\% amino-acid identity (pink).
However, the shifts between H3 and H1 are usually less than the differences between either HA and HIV Env (green).
}
\end{SCfigure*}

\subsection*{Properties associated with the shifts in amino-acid preferences between H3 and H1 HAs}
What features distinguish the sites with shifted amino-acid preferences between H3 and H1 HAs?
The sites of large shifts do not obviously localize to one specific region of HA's structure (Figure~\ref{fig:RMSD_heatmap}A).
However, at the domain level, sites in HA's stalk tend to have smaller shifts than sites in HA's globular head (Figure~\ref{fig:RMSD_heatmap}B). 
The HA stalk domain is also more conserved in sequence~\cite{nobusawa1991comparison,hai2012influenza,mallajosyula2014influenza}, suggesting that conservation of amino-acid sequence tends to be correlated with conservation of amino-acid preferences.
Consistent with this idea, sites that are absolutely conserved across all 18 HA subtypes are significantly less shifted than sites that are variable across HA subtypes (Figure~\ref{fig:RMSD_heatmap}B).
Presumably these sites are under consistent functional constraint across all HAs.

Despite their high sequence divergence, H1 and H3 adopt very similar protein folds~\cite{ha2002h5,russell2004h1}.
However, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among different HA subtypes~\cite{ha2002h5,russell2004h1}.
Previous work has defined clades of structurally related HA subtypes~\cite{ha2002h5,russell2004h1}.
One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}A).
Sites that are conserved at different amino-acid identities in these two clades tend to have exceptionally large shifts in amino acid preferences (Figure~\ref{fig:RMSD_heatmap}B).
The clade containing H1 has an upward translation of the globular head relative to the clade containing H3.
This structural shift has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
Specifically, the clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
In deep mutational scanning of the H1 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
In contrast, the interhelical loop in H3 HA makes a sharper and shorter turn which is facilitated by a Gly at 75(HA2).
In the deep mutational scanning of the Perth/2009 H3 HA, site 75(HA2) prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations.
Therefore, some of the shifts in HA amino-acid preferences can be directly rationalized in terms of changes in HA structure.

@@

\section*{Discussion}
\label{sec:discussion}
We have measured the effect of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements are informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
Specifically, HA sequences on the trunk of the phylogenetic tree tend to have mutations that our experiments measure to be more beneficial for viral growth than mutations that occur in side-branch HA sequences.
The fact that our experiments can help distinguish between trunk and side-branch lineages suggests that they might inform evolutionary forecasting.
In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, {\L}uksza and L{\"a}ssig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
Our work suggests that deep mutational scanning can provide such data.

>>>>

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,d,e]{Juhye M. Lee}
\author[b,f]{John Huddleston} 
\author[a,d,e]{Michael B. Doud}
\author[a,f]{Kathryn A. Hooper}
\author[b,c,1]{Trevor Bedford}
\author[a,c,d,1]{Jesse D. Bloom}

@@

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{J.M.L. performed the experiments; J.M.L. and J.H. analyzed data with input from T.B. and J.D.B.; M.B.D. contributed essential reagents and expertise; K.A.H. contributed essential reagents; J.M.L. and J.D.B. wrote the paper with input from J.H. and T.B.; all authors designed and planned the project}
\authordeclaration{The authors declare no conflict of interest.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed: trevor@bedford.io, jbloom@fredhutch.org}

@@

Two main characteristics distinguish the evolutionarily successful trunk from its competitors: greater antigenic change, and efficient viral growth and transmission.
In principle, experiments could be informative for identifying how mutations affect these features.
Most work on influenza evolution to date has utilized experimental data primarily to assess the antigenicity of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction,koel2013substitutions,chambers2015identification,li2016selection}.
While the antigenic features of a virus certainly contribute to its success, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
Specifically, due to influenza virus's high mutation rate~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and lack of intra-segment recombination~\cite{boni2008homologous}, deleterious mutations become linked to beneficial ones.
The resulting accumulation of deleterious mutations can affect non-antigenic properties that are central to viral fitness~\cite{luksza2014predictive}.
However, there are no large-scale quantitative characterizations of how mutations to H3N2 HA affect non-antigenic properties crucial to viral fitness.

It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of large numbers of mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high}.
However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
Here we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
We then examine whether these measurements are useful for distinguishing between H3N2 trunk and side-branch lineages, and find that mutations on the trunk are measured to be more favorable than those on the side branches, both at epitope and non-epitope sites.
However, the experiments are useful for assessing the effects of mutations only in closely related viral strains, since we show that mutations often have disparate effects on H3 and H1 HAs.
Our work highlights the potential for using high-throughput experimental measurements of mutational effects to inform evolutionary forecasting of human seasonal influenza virus.

@@

We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates. 
We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
Because there are 566 residues in HA, there are $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
The biological replicates were well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
Replicate 1 exhibited the weakest correlation with other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity during viral passage contributes most of the experimental noise.

\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
This result indicates that our experiments authentically capture some of the constraints on HA evolution. 
The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for by the constraints measured in the deep mutational scanning.
The ExpCM stringency parameter~\cite{hilton2017phydms} is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.

@@

Examination of Figure~\ref{fig:logoplot} reveals that the experimentally measured amino-acid preferences generally agree with existing knowledge about HA's structure and function.
For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} possess high preference for cysteine.
At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such as Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change,yang2015structure}.
A positively charged amino acid at site 329 is important for cleavage of the HA0 precursor into the mature form~\cite{stech2005new}, and this site strongly prefers arginine.
However, a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine. 
This codon is part of the signal peptide and is cleaved from the mature HA protein.
One possible reason that our experiments do not show a strong preference for methionine at this site could be alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.
% No upstream or downstream "ATG"'s are present. I did find an in-frame "ATC" one codon upstream of the canonical start site (in the non-coding region), and interestingly, all Perth/2009 HA sequences deposited in GenBank have an "ATT" in this non-coding position, including Seema's Perth HA.
% I6M(HA2) mutant slightly increases the fusion pH (Daniels 1985 Cell)
% We see high preference for T,S,Q over the wildtype Ile at site 226. Although Q226 in an H3 background is characterized as having preference for avian-type receptors, most of these studies were done in older H3 strains such as Aichi. Wu 2017 Cell Host Microbe has shown extensive epistasis in the RBS, specifically in the 220-loop, so perhaps the amino acids tolerated in this loop have shifted over time. Indeed the wildtype amino acids at 225 and 226 are different in Perth compared to HK68 and Aichi.

@@

\subsection*{There is less difference in mutational tolerance between the HA head and stalk domains for H3 than for H1}
Our experiments measure which amino acids are tolerated at each HA site under selection for protein function.
We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we quantify as the Shannon entropy of the re-scaled preferences.
In prior mutational studies of H1 HAs, the stalk domain was found to be substantially less mutationally tolerant than the globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,heaton2013genome}.

We performed a similar analysis using the current data for the Perth/2009 H3 HA.
Surprisingly, there was much less contrast in mutational tolerance between the stalk and head domains for the H3 HA than for the H1 (Figure~\ref{fig:mut_tolerance}).
For instance, in the H3 HA, the short helix A in the stalk is very mutationally tolerant.
Interestingly, there are more reports of readily selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.

@@

To investigate whether our experiments can aid in distinguishing trunk and side-branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the H3N2 phylogeny (Figure~\ref{fig:H3N2_phylogeny}).
The mutations on the trunk of the tree were consistently more beneficial for viral growth according to our experimental measurements (Figure~\ref{fig:trunkvssidebranch}A).
This difference in the effects of mutations between the trunk and side-branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
Some influenza sequences are derived from egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{mcwhite2016sequence}.
These mutations mostly occur on the terminal side branches that lead to tip nodes on the phylogenetic tree.
We therefore re-calculated the statistics separating side branches to tip and internal nodes, and again found that the difference between the trunk and both sets of side branches was statistically significant (Figure~\ref{fig:trunkvssidebranch}B).
Therefore, strains with mutations that we experimentally measure to be favorable for viral growth tend to do better in nature than strains with mutations that we measure to be less favorable.

@@

Mutations on the trunk are scored as significantly more favorable in our experiments than side-branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}A).
Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and at sites that are also under immune pressure.
The fact that our experiments can discriminate between the trunk and side branches at epitope sites despite only assaying for viral growth underscores the fact that mutations at these sites still have non-antigenic functional consequences~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.

However, there are differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
The per-site sequence preference at epitope sites increases over time, whereas at non-epitope sites it remains relatively constant  (Figure~\ref{fig:sequence_preference}B).
This difference is unlikely to be explained purely by differences in the number of mutations at these two sets of sites: there have been $\sim$4-fold more total trunk mutations at epitope than non-epitope sites (109 versus 28), but the large increase in sequence preference for the epitope sites is $\gg$4-fold larger than the nearly imperceptible increase for non-epitope sites.
The fact that only epitope sites exhibit an increase in sequence preference over time can be rationalized in terms of prior knowledge about protein evolution.
As proteins evolve, there can be epistatic shifts in the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some epitope sites in H3 HA~\cite{wu2017diversity}.
It is also known that immune pressure can increase epistasis among substitutions~\cite{gong2014epistatically}.
We hypothesize that immune pressure increases the accumulation of epistatic interactions involving substitutions at epitope sites.
Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA (Figure~\ref{fig:sequence_preference}B).
Such a trend is not apparent for non-epitope sites, perhaps because they are under less selection to fix adaptive mutations.

@@

An obvious hypothesis for why the H1 deep mutational scanning is not informative about the evolutionary fate of H3 strains is that the effect of the same mutation is often different between these two HA subtypes.
To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2017mapping}.
However, the H1 and H3 HAs are far more diverged, with only 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
One simple way to investigate the extent of shifts in amino-acid preferences is to correlate measurements from independent deep mutational scanning replicates on H1 and H3 HAs.
Figure~\ref{fig:distance_distribution}B shows that replicate measurements on the same HA variant are more correlated than those on different HA variants.

@@

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=12cm]{figs/distance_distribution/distance_distribution.pdf}
\caption{\label{fig:distance_distribution}
{\bf There are substantial differences in the effects of mutations between H1 and H3 HAs.}
(A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled. 
These HAs have 42\% amino-acid identity.
(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study  and the three replicates in prior deep mutational scanning of an H1 HA~\cite{doud2016accurate}.
Comparisons between H3 replicates are in purple, those between H1 replicates are in brown, and those across H1 and H3 replicates are in gray. 
$R$ indicates the Pearson correlation coefficient.
(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2017mapping}, and plotted the distribution of shifts for all sites.
The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
The shifts are also much larger than those previously observed between two variants of HIV Env that share 86\% amino-acid identity (pink).
However, the shifts between H3 and H1 are usually less than the differences between HA and HIV Env (green).
}
\end{SCfigure*}

\subsection*{Properties associated with the shifts in amino-acid preferences between H3 and H1 HAs}
What features distinguish the sites with shifted amino-acid preferences between H3 and H1 HAs?
The sites of large shifts do not obviously localize to one specific region of HA's structure (Figure~\ref{fig:RMSD_heatmap}A).
However, at the domain level, sites in HA's stalk tend to have smaller shifts than sites in HA's globular head (Figure~\ref{fig:RMSD_heatmap}B). 
The HA stalk domain is also more conserved in sequence~\cite{nobusawa1991comparison}, suggesting that conservation of amino-acid sequence is correlated with conservation of amino-acid preferences.
Consistent with this idea, sites that are absolutely conserved across all 18 HA subtypes are significantly less shifted than sites that are variable across HA subtypes (Figure~\ref{fig:RMSD_heatmap}B).
Presumably these sites are under consistent functional constraint across all HAs.

Despite their high sequence divergence, H1 and H3 adopt very similar protein folds~\cite{ha2002h5,russell2004h1}.
However, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among different HA subtypes~\cite{ha2002h5,russell2004h1}.
Previous work has defined clades of structurally related HA subtypes~\cite{ha2002h5,russell2004h1}.
One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}A).
Sites that are conserved at different amino-acid identities in these two clades tend to have exceptionally large shifts in amino acid preferences (Figure~\ref{fig:RMSD_heatmap}B).
The clade containing H1 has an upward translation of the globular head relative to the clade containing H3.
This structural shift has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
Specifically, the clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
In deep mutational scanning of the H1 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
In contrast, the interhelical loop in H3 HA makes a sharper and shorter turn which is facilitated by a Gly at 75(HA2).
In the deep mutational scanning of the Perth/2009 H3 HA, site 75(HA2) prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations.
Therefore, some of the shifts in HA amino-acid preferences can be rationalized in terms of changes in HA structure.

@@

\section*{Discussion}
\label{sec:discussion}
We have measured the effects of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements are informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
Specifically, HA sequences on the trunk of the phylogenetic tree tend to have mutations that our experiments measure to be more beneficial for viral growth than mutations that occur in side-branch HA sequences.
The fact that our experiments can help distinguish between trunk and side-branch lineages suggests that they might inform evolutionary forecasting.
In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, {\L}uksza and L{\"a}ssig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
Our work suggests that deep mutational scanning can provide such data.
